---
title: "Homework1"
author: "Zhikai Yang"
date: "02-16-2020"
output: html_notebook
---


```{r setup}
# , include=TRUE, warning=FALSE, echo=TRUE, error=FALSE
knitr::opts_knit$set(root.dir=normalizePath('../../'))
knitr::opts_chunk$set(warning=FALSE, message=FALSE, error=FALSE, echo=TRUE)
```


# NGS data simulation using `wgsim`

```
Usage:   wgsim [options] <in.ref.fa> <out.read1.fq> <out.read2.fq>

Options: -e FLOAT      base error rate [0.020]
         -d INT        outer distance between the two ends [500]
         -s INT        standard deviation [50]
         -N INT        number of read pairs [1000000]
         -1 INT        length of the first read [70]
         -2 INT        length of the second read [70]
         -r FLOAT      rate of mutations [0.0010]
         -R FLOAT      fraction of indels [0.15]
         -X FLOAT      probability an indel is extended [0.30]
         -S INT        seed for random generator [-1]
         -A FLOAT      disgard if the fraction of ambiguous 
                       bases higher than FLOAT [0.05]
         -h            haplotype mode
```


---

# Download Reference from EnsemblPlants

Maize [reference genome](https://plants.ensembl.org/Zea_mays/Info/Index)

#### Change to `largedata\hm1` folder:
```{bash, eval=FALSE}
cd largedata
mkdir hm1
cd hm1

# request nodes
srun --qos=short --nodes=1 --licenses=common --ntasks=4 --mem 32G --time 6:00:00 --pty bash
```


#### Then use `wget` to download the reference genome:

```{bash, eval=FALSE}
wget ftp://ftp.ensemblgenomes.org/pub/plants/release-46/fasta/zea_mays/dna/Zea_mays.B73_RefGen_v4.dna.chromosome.Mt.fa.gz

### then unzip it
gunzip Zea_mays.B73_RefGen_v4.dna.chromosome.Mt.fa.gz
```

---



- Reference (about 500k)
  - `Zea_mays.B73_RefGen_v4.dna.chromosome.Mt.fa`
- 20x coverage
  - `N 5000`
- PE 100bp
  - `-1 100 -2 100`
- Only SNP no Indel
  - `-R 0 -X 0`
- Mutation rate is low 
  - `-r 0.01`

---

# NGS data simulation using `wgsim`

## simulate 20 individals

```{bash, eval=FALSE}
for i in {1..20}
do
   wgsim Zea_mays.B73_RefGen_v4.dna.chromosome.Mt.fa -e 0 -d 500 -N 50000 -1 100 -2 100 -r 0.1  -R 0 -X 0 l$i.read1.fq l$i.read2.fq
done
```

--


#### check how many reads

```{bash, eval=FALSE}
wc -l l1.read1.fq 
# suppose to be 200,000 lines = 50,000 reads
```

---

# A procedure to calculate $\theta$ and $F_{ST}$ values

### 1. Align the NGS reads to the reference genome
  - [bwa](https://github.com/lh3/bwa)
  - [samtools](https://github.com/samtools/samtools)


### 2. Calculate SFS 
  - [ANGSD](http://www.popgen.dk/angsd/index.php/ANGSD)

### 3. Calculate the thetas and Fst for each site
  - ANGSD

---
# A procedure to calculate $\theta$ and $F_{ST}$ values

### 1. Align the NGS reads to the reference genome

```{bash, eval=FALSE}
module load bwa samtools
# index the reference genome
bwa index Zea_mays.B73_RefGen_v4.dna.chromosome.Mt.fa
# using bwa mem to align the reads to the reference genome 
# => samtools to convert into bam file
#bwa mem Zea_mays.B73_RefGen_v4.dna.chromosome.Mt.fa l1.read1.fq l1.read2.fq | samtools view -bSh - > l1.bam
```

#### Do alignment for 20 individuals using bash loop:

```{bash, eval=FALSE}
# alignment
for i in {1..20}; do bwa mem Zea_mays.B73_RefGen_v4.dna.chromosome.Mt.fa l$i.read1.fq l$i.read2.fq | samtools view -bSh - > l$i.bam; done
# sort
for i in *.bam; do samtools sort $i -o sorted_$i; done
# index them
for i in sorted*.bam; do samtools index $i; done
```


#### Check mapping statistics

```{bash, eval=FALSE}
samtools flagstat sorted_l1.bam
```


---

# A procedure to calculate $\theta$ values

### 2. Calculate SFS using `ANGSD`

#### Install ANGSD first

```{bash, eval=FALSE}
cd ~/bin/ # if you don't have one, do `mkdir bin`
git clone https://github.com/samtools/htslib.git
git clone https://github.com/ANGSD/angsd.git 
cd htslib; make; 
cd ../angsd;
make HTSSRC=../htslib
```

#### run angsd

```{bash, eval=FALSE}
# write the bam files to a txt file
mkdir bam_files
mv sorted*.bam bam_files
cd bam_files
ls sorted*.bam > bam.txt
```

---

# A procedure to calculate $\theta$ values


```{bash, eval=FALSE}
angsd -bam bam.txt -doSaf 1 -anc ../Zea_mays.B73_RefGen_v4.dna.chromosome.Mt.fa -GL 1  -out out 
# use realSFS to calculate sfs
realSFS out.saf.idx > out.sfs
```

#### Copy the result to `cache/` folder

```{bash, eval=FALSE}
## cp sfs to the cache/ folder
cp out.sfs ../../../cache/
```



### 3. Calculate the thetas for each site

The output from the above command are two files out.thetas.gz and out.thetas.idx. A formal description of these files can be found in the doc/formats.pdf in the angsd package. It is possible to extract the logscale persite thetas using the ./thetaStat print program.

```{bash, eval=FALSE}
angsd -bam bam.txt -out out -doThetas 1 -doSaf 1 -pest out.sfs -anc ../Zea_mays.B73_RefGen_v4.dna.chromosome.Mt.fa -GL 1
thetaStat print out.thetas.idx > theta.txt
## cp theta to the cache/ folder
cp theta.txt ../../../cache/
```

---

# A procedure to calculate $F_{ST}$

#### Create two list bam files


#### Two population Fst
```{bash, eval=FALSE}
# first calculate per pop saf for each populatoin
angsd -b pop1.txt -anc ../Zea_mays.B73_RefGen_v4.dna.chromosome.Mt.fa -out pop1 -dosaf 1 -gl 1
angsd -b pop2.txt -anc ../Zea_mays.B73_RefGen_v4.dna.chromosome.Mt.fa -out pop2 -dosaf 1 -gl 1
# calculate the 2dsfs prior
realSFS pop1.saf.idx pop2.saf.idx > pop1.pop2.ml
# prepare the fst for easy window analysis etc
realSFS fst index pop1.saf.idx pop2.saf.idx -sfs pop1.pop2.ml -fstout out
# get the global estimate
realSFS fst stats out.fst.idx 
# below is not tested that much, but seems to work
realSFS fst stats2 out.fst.idx -win 500 -step 100 > fst_win.txt

```

#### Copy the result to `cache/` folder

```{bash, eval=FALSE}
## cp sfs to the cache/ folder
cp fst_win.txt ../../../cache/
```

---

# Visualize the results

In local computer, using `R`:

#### Barplot for SFS

```{r, eval=FALSE}
library(data.table)
s <- scan('cache/out.sfs')
s <- s[-c(1,length(s))]
s <- s/sum(s)
barplot(s,names=1:length(s), main='SFS')
```

#### Histgram distribution of the theta values

```{r, eval=FALSE}
theta <- fread("cache/theta.txt", data.table=FALSE)
hist(theta$Pairwise, main = "Pairwise Theta Distribution", xlab = "Log10(Pairwise_Theta)")
```

#### Scatter plot of the Fst values


```{r, eval=FALSE}
fst <- read.table("cache/fst_win.txt", skip=1, header=FALSE)
names(fst)[c(3,5)] <- c("midp", "fst")
plot(fst$midp, fst$fst, xlab="Physical position", ylab="Fst", col="#5f9ea0", pch=16)
```

---

# General feature format (GFF) from EnsemblPlants

Maize [reference genome](https://plants.ensembl.org/Zea_mays/Info/Index)

change to `largedata\hm1` folder:

```{bash, eval=FALSE}
wget ftp://ftp.ensemblgenomes.org/pub/plants/release-46/fasta/zea_mays/dna/Zea_mays.B73_RefGen_v4.dna.chromosome.Mt.fa.gz

### then unzip it
gunzip Zea_mays.B73_RefGen_v4.dna.chromosome.Mt.fa.gz
```

Similarly, we will download and unzip the [Mt GFF3](ftp://ftp.ensemblgenomes.org/pub/plants/release-46/gff3/zea_mays/Zea_mays.B73_RefGen_v4.46.chromosome.Mt.gff3.gz) file

--

#### Use R to process the GFF3 file

```{r, eval=FALSE}
# install.package("data.table")
library("data.table")


###only works under Unix system not windows
## grep -v means select lines that not matching any of the specified patterns
gff <- fread(cmd='grep -v "#" largedata/hm1/Zea_mays.B73_RefGen_v4.46.chromosome.Mt.gff3', header=FALSE, data.table=FALSE)

fwrite(gff, "largedata/hm1/gff.txt", sep="\t", row.names=FALSE, quote=FALSE)

###the output from Unix now can work for windows
gff <- fread("largedata/hm1/gff.txt",  header=T, data.table=FALSE)

```

---

# General feature format (GFF) version 3

```
  V1      V2         V3   V4     V5 V6 V7 V8
1 Mt Gramene chromosome    1 569630  .  .  .
2 Mt ensembl       gene 6391   6738  .  +  .
3 Mt    NCBI       mRNA 6391   6738  .  +  .
4 Mt    NCBI       exon 6391   6738  .  +  .
5 Mt    NCBI        CDS 6391   6738  .  +  0
6 Mt ensembl       gene 6951   8285  .  +  .
                  V9
1   ID=chromosome:Mt;Alias=AY506529.1,NC_007982.1;Is_circular=true
2   ID=gene:ZeamMp002;biotype=protein_coding;description=orf115-a1;
3   ID=transcript:ZeamMp002;Parent=gene:ZeamMp002;
4   Parent=transcript:ZeamMp002;Name=ZeamMp002.exon1;constitutive=1;ensembl_end_phase=0;
5   ID=CDS:ZeamMp002;Parent=transcript:ZeamMp002;
6   ID=gene:ZeamMp003;biotype=protein_coding;description=orf444
```

---

# General feature format (GFF) version 3

```
  V1      V2         V3   V4     V5 V6 V7 V8
1 Mt Gramene chromosome    1 569630  .  .  .
2 Mt ensembl       gene 6391   6738  .  +  .
                  V9
1   ID=chromosome:Mt;Alias=AY506529.1,NC_007982.1;Is_circular=true
2   ID=gene:ZeamMp002;biotype=protein_coding;description=orf115-a1;
```
--------------

- 1	__sequence__:	The name of the sequence where the feature is located.

- 2	__source__:	Keyword identifying the source of the feature, like a program (e.g. RepeatMasker) or an organization (like ensembl).

- 3	__feature__:	The feature type name, like "gene" or "exon". 
  - In a well structured GFF file, all the children features always follow their parents in a single block (so all exons of a transcript are put after their parent "transcript" feature line and before any other parent transcript line).
  
- 4	__start__:	Genomic start of the feature, with a 1-base offset. 
  - This is in contrast with other 0-offset half-open sequence formats, like [BED]().
  

---
# General feature format (GFF) version 3

```
  V1      V2         V3   V4     V5 V6 V7 V8
1 Mt Gramene chromosome    1 569630  .  .  .
2 Mt ensembl       gene 6391   6738  .  +  .
                  V9
1   ID=chromosome:Mt;Alias=AY506529.1,NC_007982.1;Is_circular=true
2   ID=gene:ZeamMp002;biotype=protein_coding;description=orf115-a1;
```

--------------

- 5	__end__:	Genomic end of the feature, with a 1-base offset. 
  - This is the same end coordinate as it is in 0-offset half-open sequence formats, like BED.

- 6	__score__:	Numeric value that generally indicates the confidence of the source in the annotated feature. 
  - A value of "." (a dot) is used to define a null value.
  
- 7	__strand__:	Single character that indicates the strand of the feature. 
  - it can assume the values of "+" (positive, or 5' -> 3'), 
  - "-", (negative, or 3' -> 5'), "." (undetermined).

---
# General feature format (GFF) version 3

```
  V1      V2         V3   V4     V5 V6 V7 V8
1 Mt Gramene chromosome    1 569630  .  .  .
2 Mt ensembl       gene 6391   6738  .  +  .
                  V9
1   ID=chromosome:Mt;Alias=AY506529.1,NC_007982.1;Is_circular=true
2   ID=gene:ZeamMp002;biotype=protein_coding;description=orf115-a1;
```

--------------

- 8	__phase__:	phase of CDS (__means CoDing Sequence__) features. 
  - The phase indicates where the feature begins with reference to the reading frame. 
  - The phase is one of the integers 0, 1, or 2, indicating the number of bases that should be removed from the beginning of this feature to reach the first base of the next codon.
  
- 9	__attributes__:	All the other information pertaining to this feature. 
  - The format, structure and content of this field is the one which varies the most between the three competing file formats.

---

# Work with GFF

```{r, eval=FALSE}
names(gff) <- c("seq", "source", "feature", "start", "end", "score", "strand", "phase", "att")
table(gff$feature)
```

### Get genes and upstream and downstream 5kb regions

```{r, eval=FALSE}
g <- subset(gff, feature %in% "gene")
g$geneid <- gsub(".*gene:|;biotype.*", "", g$att)

### + strand
gp <- subset(g, strand %in% "+") 
# nrow(gp) 75

### get the 5k upstream of the + strand gene model
gp_up <- gp
gp_up$end <- gp_up$start - 1
gp_up$start <- gp_up$end - 5000 

### get the 5k downstream of the + strand gene model
gp_down <- gp
gp_down$start <- gp_down$end + 1
gp_down$end <- gp_down$start + 5000 
```

---


### Get genes and upstream and downstream 5kb regions

```{r echo=FALSE, eval=FALSE}
g <- subset(gff, feature %in% "gene")
g$geneid <- gsub(".*gene:|;biotype.*", "", g$att)

### - strand
gm <- subset(g, strand %in% "-") 
# nrow(gm) 82

### get the 5k upstream of the + strand gene model
gm_up <- gm
gm_up$start <- gm_up$end + 1
gm_up$end <- gm_up$start + 5000 

### get the 5k downstream of the + strand gene model
gm_down <- gm
gm_down$end <- gm_down$start - 1
gm_down$start <- gm_down$end - 5000 

gup <- rbind(gp_up, gm_up)
fwrite(gup, "cache/mt_gene_up5k.txt", sep="\t", row.names = FALSE, quote=FALSE)

gdown <- rbind(gp_down, gm_down)
fwrite(gdown, "cache/mt_gene_down5k.txt", sep="\t", row.names = FALSE, quote=FALSE)
```


```{r, eval=FALSE}

fwrite(g, "cache/mt_gene.txt", sep="\t", row.names = FALSE, quote=FALSE)
```

---

## Intepret the theta results


```{r, eval=FALSE}
library("data.table")
library("GenomicRanges")
library("plyr")


theta <- fread("cache/theta.txt", data.table=FALSE)
names(theta)[1] <- "seq"

up5k <- read.table("cache/mt_gene_up5k.txt", header=TRUE)

### define the subject file for theta values
grc <- with(theta, GRanges(seqnames=seq, IRanges(start=Pos, end=Pos)))

### define the query file for genomic feature
grf <- with(up5k, GRanges(seqnames=seq, IRanges(start=start, end=end), geneid=geneid))
    
### find overlaps between the two
tb <- findOverlaps(query=grf, subject=grc)
tb <- as.matrix(tb)
    
out1 <- as.data.frame(grf[tb[,1]])
out2 <- as.data.frame(grc[tb[,2]])
### for each genomic feature, find the sites with non-missing data
out <- cbind(out1, out2[, "start"]) 
names(out)[ncol(out)] <- "pos"


```

---

## Intepret the theta results


```{r, eval=FALSE}
#define unique identifier and merge with the thetas
out$uid <- paste(out$seqnames, out$pos, sep="_")
theta$uid <- paste(theta$seq, theta$Pos, sep="_")

df <- merge(out, theta[, c(-1, -2)], by="uid")
# for each upstream 5k region, how many theta values

mx <- ddply(df, .(geneid), summarise,
            Pairwise = mean(Pairwise, na.rm=TRUE),
            thetaH = mean(thetaH, na.rm=TRUE),
            nsites = length(uid))

```

---

## Intepret the theta results

Copy and paste everything above, and pack it into an `R` function:

```{r, eval=FALSE}
get_mean_theta <- function(gf_file="cache/mt_gene_up5k.txt"){
  # gf_file: gene feature file [chr, ="cache/mt_gene_up5k.txt"]
  
  theta <- fread("cache/theta.txt", data.table=FALSE)
  names(theta)[1] <- "seq"

  up5k <- read.table(gf_file, header=TRUE)

  ### define the subject file for theta values
  grc <- with(theta, GRanges(seqnames=seq, IRanges(start=Pos, end=Pos)))

  ### define the query file for genomic feature
  grf <- with(up5k, GRanges(seqnames=seq, IRanges(start=start, end=end), geneid=geneid))
    
  ### find overlaps between the two
  tb <- findOverlaps(query=grf, subject=grc)
  tb <- as.matrix(tb)
    
  out1 <- as.data.frame(grf[tb[,1]])
  out2 <- as.data.frame(grc[tb[,2]])
  ### for each genomic feature, find the sites with non-missing data
  out <- cbind(out1, out2[, "start"]) 
  names(out)[ncol(out)] <- "pos"

  #define unique identifier and merge with the thetas
  out$uid <- paste(out$seqnames, out$pos, sep="_")
  theta$uid <- paste(theta$seq, theta$Pos, sep="_")

  df <- merge(out, theta[, c(-1, -2)], by="uid")
  # for each upstream 5k region, how many theta values

  mx <- ddply(df, .(geneid), summarise,
            Pairwise = mean(Pairwise, na.rm=TRUE),
            thetaH = mean(thetaH, na.rm=TRUE),
            nsites = length(uid))
  return(mx)
}
```



---

## Plot the results

Run the customized R function

```{r, eval=FALSE}
### apply the function
up5k <- get_mean_theta(gf_file="cache/mt_gene_up5k.txt")
down5k <- get_mean_theta(gf_file="cache/mt_gene_down5k.txt")
gene <- get_mean_theta(gf_file="cache/mt_gene.txt")

```

And then plot the results:

```{r, eval=FALSE}
library("ggplot2")

up5k$feature <- "up 5k"
down5k$feature <- "down 5k"
gene$feature <- "genic"
res <- rbind(up5k, down5k)
res_nongenic <- res
res_nongenic$feature <- "intergenic"
res_t <- rbind(res, gene, res_nongenic)


ggplot(res_t, aes(x=feature, y=Pairwise, fill=feature)) + 
  geom_violin(trim=FALSE)+
  labs(title="Theta value", x="", y = "Log10 (theta)")+
  geom_boxplot(width=0.1, fill="white")+
  scale_fill_brewer(palette="Blues") + 
  theme_classic()

```


## Intepret the Fst results

```{r, eval=FALSE}
get_mean_Fst <- function(gf_file="cache/mt_gene_up5k.txt"){
  # gf_file: gene feature file [chr, ="cache/mt_gene_up5k.txt"]
  
  fst <- fread("cache/fst_win.txt", data.table=FALSE)
  names(fst)[2] <- "seq"

  up5k <- read.table(gf_file, header=TRUE)

  ### define the subject file for theta values
  grc_f <- with(fst, GRanges(seqnames=seq, IRanges(start=chr, end=chr)))

  ### define the query file for genomic feature
  grf <- with(up5k, GRanges(seqnames=seq, IRanges(start=start, end=end), geneid=geneid))
    
  ### find overlaps between the two
  tb_f <- findOverlaps(query=grf, subject=grc_f)
  tb_f <- as.matrix(tb_f)
    
  out1_f <- as.data.frame(grf[tb_f[,1]])
  out2_f <- as.data.frame(grc[tb_f[,2]])
  ### for each genomic feature, find the sites with non-missing data
  out_f <- cbind(out1_f, out2_f[, "start"]) 
  names(out_f)[ncol(out_f)] <- "pos"

  #define unique identifier and merge with the thetas
  out_f$uid <- paste(out_f$seqnames, out_f$pos, sep="_")
  fst$uid <- paste(fst$seq, fst$chr, sep="_")

  df_f <- merge(out_f, fst[, c(-1, -2, -3, -4)], by="uid")
  # for each upstream 5k region, how many theta values
  names(df_f)[ncol(df_f)] <- "fst"
  
  mx_f <- ddply(df_f, .(geneid), summarise,
            Fst = mean(fst, na.rm=TRUE),
            nsites = length(uid))
  return(mx_f)
}
```
## Plot the results

Run the customized R function

```{r, eval=FALSE}
### apply the function
up5k_f <- get_mean_Fst(gf_file="cache/mt_gene_up5k.txt")
down5k_f <- get_mean_Fst(gf_file="cache/mt_gene_down5k.txt")
gene_f <- get_mean_Fst(gf_file="cache/mt_gene.txt")

```

And then plot the results:

```{r, eval=FALSE}
library("ggplot2")

up5k_f$feature <- "up 5k"
down5k_f$feature <- "down 5k"
gene_f$feature <- "genic"
res_f <- rbind(up5k_f, down5k_f)
res_nongenic_f <- res_f
res_nongenic_f$feature <- "intergenic"
res_t_f <- rbind(res_f, gene_f, res_nongenic_f)


ggplot(res_t_f, aes(x=feature, y=Fst, fill=feature)) + 
  geom_violin(trim=FALSE)+
  labs(title="Fst distribution", x="", y = "Fst")+
  geom_boxplot(width=0.1, fill="white")+
  scale_fill_brewer(palette="Reds") + 
  theme_classic()

```
